RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-25 doc/paper.tex
Section 3.3.2 i 3.3.3 upodate

Perturbations allow the algorithm to escape from local optima. As previously described, perturbations are performed in two occasions: ($i$) during the exhaustive examination of neighboring solutions around the starting solution, ($ii$)during selection of the next starting solution, a non-exhaustive case.  
In both cases, the same Algorithm~\ref{alg:pert} is used.

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local perturbations (1-perturbations) of solution $s$ -- $s_{perturbations}$
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$ \label{line:pertInit}
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$n \in s_{subtrees}$} \label{line:forSSStart}
			\State $n_{perturbations} \gets$ \texttt{All1PerturbationsAroundNode}($s$, $n$)
			\State $s_{perturbations} \gets s_{perturbations} \cup n_{perturbations}$	
		\EndFor \label{line:forSSEnd}
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of all 1-perturbations for a given solution.}
	\label{alg:pert}
\end{algorithm}

@@

\begin{algorithm}	
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$, node $n$\\
	\hspace*{\algorithmicindent} \textbf{Output}: 1-perturbations of solution $s$ around node $n$
	\begin{algorithmic}[1]
		\Procedure{All1PerturbationsAroundNode}{$s$, $n$}
		\State $n_{perturbations} \gets \emptyset$
		\If{$n = s$}
		\State $n_{perturbations} \gets n_{perturbations} \cup$ \texttt{NodeChanges}($n$)
		\EndIf
		\If{$n.arity \geq 1$}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.left$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.left$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\If{$n.arity = 2$}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.right$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.right$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\State \Return $n_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations for a given solution around given node.}
	\label{alg:pertNode}
\end{algorithm}

@@

\emph{Case 1}. 
Observed node $n$ is the whole tree $s$. Following on the the previous exemplary expression tree, this means that multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be replaced by addition, which forms a perturbation $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Node $n$ has arity of at least 1. This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $n=x+y^3$ the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is replaced by division). 
Another example would be the case of unary operation, e.g. when $n=\sqrt{y}$. In that case, some of the possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

\emph{Case 3}. 
Subtree $ss$ is binary operation, meaning the right subtree must exist. 
The analogous idea is applied as in \emph{Case 2}.

@@

\begin{enumerate}
	\item Any node to any of its subtrees (excluding itself). For example, if $(x+y^3)$ is changed to $x$, the perturbation is $(x+y^3)\cdot x/\sqrt{y} \rightarrow x\cdot x/\sqrt{y}$. 
	\item Variable to another variable or constant to variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (y+y^3)\cdot x/\sqrt{y}$.
	\item Variable to unary operation applied to that variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot \ln{x}/\sqrt{y}$.
	\item Unary operation to another unary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sin{y}$.
	\item Binary operation to another binary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot (x + \sqrt{y})$. 
	\item Variable or constant enter the binary operation with arbitrary variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+x/y^3)\cdot x/\sqrt{y}$. 
\end{enumerate} 
\fxnote{Dati kompleksnost velicine skupa perturbacija u odnosu na velicinu drveta i dopustene promjene}

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: perturbation $p$, sample training dataset $D_{tr}'$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local optimum $bp$ in the vicinity of perturbation $p$
	\begin{algorithmic}[1] 
		\Procedure{LocalSearch}{$p$, $D_{tr}'$}
		\State $bp, bp_{fit} \gets p, $\texttt{Fitness}$(p, D_{tr}')$ 
		\State $improved \gets true$
		\While{$improved$}
			\State $improved \gets false$
			\State $bp_{candidates} \gets \emptyset$
			\State $bp_{subtrees} \gets $\texttt{SubTrees}($bp$)
			\For{$n \in bp_{subtrees}$}
				\State $n_{candidates} \gets $ \texttt{All1PerturbationsAroundNode}($bp$, $n$)
				\For{$new \in n_{candidates}$}
					\State $new \gets$ \texttt{FitOLS}($new$, $D_{tr}'$)
					\State $new_{fit} \gets$ \texttt{Fitness}($new$, $D_{tr}'$)
					\If{$new_{fit} < bp_{fit}$}
						\State $bp, bp_{fit} \gets new, new_{fit}$
						\State $improved \gets true$
					\EndIf
				\EndFor
			\EndFor
		\EndWhile
		\State \Return $bp$
		\EndProcedure
	\end{algorithmic}
	\caption{Local search procedure.}
	\label{alg:ls}
\end{algorithm}

For a given perturbation $p$, local search systematically explores all 1-perturbations around $p$. It relies on the \emph{best-improvement} strategy, meaning that all 1-perturbations (for all subtrees) are considered. Before checking if the candidate solution ($new$) is better that the actual best $bp$, OLS coefficient fitting is performed.

@@

\end{itemize}
The abovementioned benchmark sets may be biased as models which describe phisical laws may have various symmetries, periodically w.r.t. some variables, internal separability on some variables etc. In that way, the search process and the search space can be biased and in-advance significantly reduced. Thus, some methods may get significant help with this regards, while some others may be in the subordinate situation. 
  In order to permit any kind of favoritism, we generate a set of randomly generated instances, called \textsc{Random}. These instances are generated according the following procedure. \fxnote{Opisati proceduru generisanja instanci + kombinaciju parametara za instance, itd.}

>>>>

Perturbations allow the algorithm to escape from local optima. As previously described, perturbations are performed in two occasions: ($i$) during the exhaustive examination of neighboring solutions around the starting solution, ($ii$) during selection of the next starting solution, a non-exhaustive case.  
In both cases, the same Algorithm~\ref{alg:pert} is used.

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local perturbations (1-perturbations) of solution $s$ -- $s_{perturbations}$
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$ \label{line:pertInit}
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$n \in s_{subtrees}$} \label{line:forSSStart}
			\State $n_{perturbations} \gets$ \texttt{All1PerturbationsAroundNode}($s$, $n$)
			\State $s_{perturbations} \gets s_{perturbations} \cup n_{perturbations}$	
		\EndFor \label{line:forSSEnd}
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of all 1-perturbations of a given solution.}
	\label{alg:pert}
\end{algorithm}

@@

\begin{algorithm}	
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$, node $n$\\
	\hspace*{\algorithmicindent} \textbf{Output}: 1-perturbations of solution $s$ around node $n$
	\begin{algorithmic}[1]
		\Procedure{All1PerturbationsAroundNode}{$s$, $n$}
		\State $n_{perturbations} \gets \emptyset$
		\If{$n = s$} \label{alg:alg4-case-1}
		\State $n_{perturbations} \gets n_{perturbations} \cup$ \texttt{NodeChanges}($n$)
		\EndIf
		\If{$n.arity \geq 1$}\label{alg:alg4-case-2}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.left$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.left$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\If{$n.arity = 2$}\label{alg:alg4-case-3}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.right$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.right$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\State \Return $n_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations of a given solution around given node.}
	\label{alg:pertNode}
\end{algorithm}

@@

\emph{Case 1}. 
Observed node $n$ is the whole tree $s$ (see line ~\ref{alg:alg4-case-1} in Algorithm~\ref{alg:pertNode}). Following on   the previous exemplary expression tree, this means that multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be replaced by addition, which forms a 1-perturbation expression (tree) $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Node $n$ has arity of at least 1 (see line ~\ref{alg:alg4-case-2} in Algorithm~\ref{alg:pertNode}). This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $n=x+y^3$ the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is replaced by division). 
Another example would be the case of unary operation, e.g. when $n=\sqrt{y}$. In that case, some of the possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

\emph{Case 3}. 
Subtree $ss$ is binary operation, meaning the right subtree must exist (Line ~\ref{alg:alg4-case-3} in Algorithm~\ref{alg:pertNode}). 
The analogous idea is applied as in \emph{Case 2}.

@@

\begin{enumerate}
	\item Any node to any of its subtrees (excluding itself). For example, if $(x+y^3)$ is changed to $x$, the perturbation is $(x+y^3)\cdot x/\sqrt{y} \rightarrow x\cdot x/\sqrt{y}$. 
	\item Variable to another variable or constant to variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (y+y^3)\cdot x/\sqrt{y}$.
	\item Variable to unary operation applied to that variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot \ln{x}/\sqrt{y}$.
	\item Unary operation to another unary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot x/\sin{y}$.
	\item Binary operation to another binary operation. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+y^3)\cdot (x + \sqrt{y})$. 
	\item Variable or constant enter the binary operation with arbitrary variable. For example,  $(x+y^3)\cdot x/\sqrt{y} \rightarrow (x+x/y^3)\cdot x/\sqrt{y}$. 
\end{enumerate} 
\fxnote{Nekako zbunjuje u Algoritmu 4, pojava NodeChanges)() i Replace(). Mislim, razumijem iz opisa sta one rade, ali nisam siguran da treba imenovati metode, ako ih barem u kontekstu opisa ne pomenemo direktno po nazivu. }

\fxnote{Dati kompleksnost velicine skupa perturbacija u odnosu na velicinu drveta i dopustene promjene}

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: perturbation $p$, sample training dataset $D_{tr}'$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local optimum $bp$ in the vicinity of perturbation $p$
	\begin{algorithmic}[1] 
		\Procedure{LocalSearch}{$p$, $D_{tr}'$}
		\State $bp, bp_{fit} \gets p,\ $\texttt{Fitness}$(p,D_{tr}')$ 
		\State $improved \gets true$
		\While{$improved$}
			\State $improved \gets false$
			\State $bp_{candidates} \gets \emptyset$
			\State $bp_{subtrees} \gets $\texttt{SubTrees}($bp$)
			\For{$n \in bp_{subtrees}$}
				\State $n_{candidates} \gets $ \texttt{All1PerturbationsAroundNode}($bp$, $n$)
				\For{$new \in n_{candidates}$}
					\State $new \gets$ \texttt{FitOLS}($new$, $D_{tr}'$)
					\State $new_{fit} \gets$ \texttt{Fitness}($new$, $D_{tr}'$)
					\If{$new_{fit} < bp_{fit}$}
						\State $bp, bp_{fit} \gets new, new_{fit}$
						\State $improved \gets true$
					\EndIf
				\EndFor
			\EndFor
		\EndWhile
		\State \Return $bp$
		\EndProcedure
	\end{algorithmic}
	\caption{Local search procedure.}
	\label{alg:ls}
\end{algorithm}

For a given perturbation $p$, local search systematically explores all 1-perturbations around $p$. It relies on the \emph{best-improvement} strategy, meaning that all 1-perturbations (for all subtrees) are considered. Before checking if the candidate solution ($new$) is better that the actual best $bp$, the OLS coefficient fitting (\texttt{FitOLS}) is performed.

@@

\end{itemize}
The above-mentioned benchmark sets may be biased as models which describe physical laws may have various symmetries, periodically w.r.t. some variables, internal separability on some variables etc. In that way, the search process and the search space can be biased and in-advance significantly reduced. Thus, some methods may get significant help with this regards, while some others may be in the subordinate situation. 
  In order to permit any kind of favoritism, we generate a set of randomly generated instances, called \textsc{Random}. These instances are generated according the following procedure. \fxnote{Opisati proceduru generisanja instanci + kombinaciju parametara za instance, itd.}
